HO unification from object language to meta language

### 2024-03-26 paper/main.tex
wip: eta and beta

\newcommand{\EqualRel}{\ensuremath{=}}
\newcommand{\nEqualRel}{\ensuremath{\new}}
\newcommand{\UnifRel}{\ensuremath{\simeq}}
\newcommand{\nUnifRel}{\ensuremath{\not\simeq}}

@@

As we mentioned in section~\ref{sec:nutshell} the compiler
replaces terms in \maybebeta and \maybebeta with fresh variables
linked to the problematic terms. Each class of problematic terms
has a dedicated link.

@@

\noindent
If the variable is assigned during a run the link is considered for
progress and possibly eliminated. This is discussed in
section~\ref{sec:eta}.\marginpar{is this the right section?}

@@

\begin{lemma}
Properties \ref{prop:correct} and
\ref{prop:complete} hold for the implementation of \Uo above
\end{lemma}
\begin{proof}[Proof sketch]
 In this setting \Ee is as strong as
\Eo on ground terms. What we have to show is that whenever two different \Fo
terms can be made equal by a substitution $\rho$ (plus the \ref{clause:beta1}
and \ref{clause:beta2} if needed) we can find this $\rho$ by finding
a $\sigma$ via \Ue{} on the corresponding \Ho terms and by decompiling it.
If we look at the \Fo{} terms, the are two interesting cases:
\begin{itemize}
\item \elpiIn{fuva X ~\Uo~ s}. In this case after \elpiIn{comp} we have
  $Y \Ue t$ that succeeds with $\sigma = \{ Y \mapsto t\}$ and
  $\sigma$ is decompiled to $\rho = \{ Y \mapsto s\}$.
\item \elpiIn{fall[fuva X|L] ~\Uo~ s}. In this case
 we have $Y_{\vec{x}} \Ue t$ that succeeds with
 $\sigma = \{ \vec{y} \vdash Y \mapsto t[\vec{x}/\vec{y}]\}$ that in turn
 is decompiled to $\rho = \{ Y \mapsto \lambda \vec{y}.s[\vec{x}/\vec{y}]\}$.
 Thanks to \ref{clause:beta1}
 $(\lambda \vec{y}.s[\vec{x}/\vec{y}])~\vec{x} \Eo s$.
\end{itemize}
Since the mapping is a bijection occur check in \Ho{} corresponds to occur
check in \Fo{}.
\end{proof}

@@

Even though the unification process explained in the previous sections is able
to solve a large number of unification problems, it remains still incomplete:
\wellb is only a subset of terms in \Ho. In order to capture all the unification
properties of \Eo, we need ad-hoc compilation strategies over those subterms
that have been defined as ``problematic''.

\subsection{Compilation}
\subsection{Progress}

@@

and aux

\otext{The following goal necessita v1 (lo scope è usato):\\ \elpiIn{X = lam x\ lam y\ Y y x, X = lam x\ f}}
\otext{The snd unif pb, we have to unif \elpiIn{lam x\ lam y\ Y x y} with \elpiIn{lam x\ f}}
\otext{It is not doable, with the same elpi var}

Invarianti:
A destra della eta abbiamo sempre un termine che comincia per $\lambda x. bla$

>>>>

\newcommand{\appsep}{\ensuremath{\textcolor{lightgray}{\cdot}}}
\newcommand{\EqualRel}{\ensuremath{=}}
\newcommand{\nEqualRel}{\ensuremath{\new}}
\newcommand{\UnifRel}{\ensuremath{\simeq}}
\newcommand{\nUnifRel}{\ensuremath{\not\simeq}}

@@

As we mentioned in section~\ref{sec:nutshell} the compiler
replaces terms in \maybeeta and \maybebeta with fresh variables
linked to the problematic terms. Each class of problematic terms
has a dedicated link.

@@

% TODO: io sposteri l'invariante sopra nella maybe-eta, perchè  
% la proprietà è vera solo per i link eta
% per i link beta, questo non è vero se sospendiamo i link
\noindent
If the variable is assigned during a run the link is considered for
progress and possibly eliminated. This is discussed in
section~\ref{sec:eta}.\marginpar{is this the right section?}

@@

\begin{lemma}
Properties \ref{prop:correct} and
\ref{prop:complete} hold for the implementation of \Uo above
\end{lemma}
\begin{proof}[Proof sketch]
 In this setting \Ee is as strong as
\Eo on ground terms. What we have to show is that whenever two different \Fo
terms can be made equal by a substitution $\rho$ (plus the \ref{clause:beta1}
and \ref{clause:beta2} if needed) we can find this $\rho$ by finding
a $\sigma$ via \Ue{} on the corresponding \Ho terms and by decompiling it.
If we look at the \Fo{} terms, the are two interesting cases:
\begin{itemize}
\item \elpiIn{fuva X ~\Uo~s}. In this case after \elpiIn{comp} we have
  $Y \Ue t$ that succeeds with $\sigma = \{ Y \mapsto t\}$ and
  $\sigma$ is decompiled to $\rho = \{ Y \mapsto s\}$.
\item \elpiIn{fapp[fuva X|L] ~\Uo~s}. In this case
 we have $Y_{\vec{x}} \Ue t$ that succeeds with
 $\sigma = \{ \vec{y} \vdash Y \mapsto t[\vec{x}/\vec{y}]\}$ that in turn
 is decompiled to $\rho = \{ Y \mapsto \lambda \vec{y}.s[\vec{x}/\vec{y}]\}$.
 Thanks to \ref{clause:beta1}
 $(\lambda \vec{y}.s[\vec{x}/\vec{y}])~\vec{x} \Eo s$.
\end{itemize}
Since the mapping is a bijection occur check in \Ho{} corresponds to occur
check in \Fo{}.
\end{proof}

@@

$\eta$-reduction is a property by which a term $t$ is of the form $\lambda x. T
x$ and $x$ is not free in $T$. In this case we case say that $T$ is equal to its
$\eta$-expanded version represented by the term $t$. Both the meta and the
object language accept terms unification up to $\eta$ equivalence, but as
sketched in the small code snippet in \cref{sec:problem-statement}, \Ue is not
able to capture this equivalence property when using terms in \Fo. The main
reason is that the unification of two terms in \Fo{} appears to be syntactical
on the constructors of these terms. As a very concise example, \Ue can't unify
the \Fo{} terms \elpiIn{lam x\ app[con "f", x]} and \elpiIn{con "f"} since the
rigid constructor \elpiIn{lam} is different from the rigid constructor
\elpiIn{con}. In conclusion, we need to adapt our \elpiIn{comp} relation in
order to address this problem.

\paragraph{Detection of \maybeeta} The main modification of the compiler to solve this
unification issue consists in identifying all the subterms of a term $t$ that
are \maybeeta. In particular, a term $t$ is a \maybeeta, if it can reduce to a
$\eta$-expansion under a certain substitution $\sigma$, the code verifying this
property is given by the following code:

The entry point is depicted by the rule~\ref{rule:maybeeta} which takes a name
$n$, a term $t$ and a list of bound variables $L$ (originally it is the
singleton containing $n$). This rule checks if $t$ is a term of the form $T
\appsep n$ (for a term $T$), together with the auxiliary predicate
\elpiIn{reducible-to} which ensures if a term $t$ can reduce to a name $n$.
\marginpar{explain better: we start from $\lambda x.F x$}
The \elpiIn{maybe-eta} predicate dispatches the calls to \elpiIn{reducible-to};
three cases should be considered: 1) $t$ is a variable $v$, then $t$ can be an
$\eta$-expansion if at least one of the terms in the scope of $v$ is a
$\maybeeta$ of $n$; 2) $t$ is a lambda-term, then we recursively call
\elpiIn{maybe-eta} on the body of $t$ under a local name $x$ which is added to
the list $L$; 3) $t$ is an application, then $t$ is an $\eta$ expansion if i)
the last arguments of $t$ can be reduced one by one to the binders in the list
$L$ (we reverse the list in rule, since, by construction, this list is built in
reversed order) and ii) none of the first arguments of the application contain a
rigid occurrence of name in $L$.

As rapidly said before, \elpiIn{reducible-to} tells if a term $t$ reduce to a
name $n$, or equivalently if $\exists \sigma, \sigma t = n$. This predicate also
takes the list of all the binders explored (this list is originally empty). A
term $t$ reduces to a name $n$ if 1) $n = t$; 2) $t$ is a variable $v$, then $t$
reduce to a $n$ if it exists an arugment in the scope of $v$ reducing to $n$ and
forall name $n'$ in $L$, there is an argument reducing to $n'$ \marginpar{Note
that this function is partial: the arg in the scope should be distinct}; 3) $t$
is a lambda abstraction, then we call recursively \elpiIn{reducible-to} on the
body of the abstraction with a new local name added to the list $L$; 4) $t$ is
an application of $n$ to a list of arguments $L'$, then all the arguments should
reduce to the respective name in the list $L$.

Finally, a name $n$ occurs rigidly in a term $t$ if $n$ occurs in a subterm $t'$
of $t$ such that $t'$ does not appear in the scope of a variable.

An example of \maybeeta detection over the bound variable $x$ is the following:
\marginpar{maybe more examples in appendix}
%
\begin{gather}
  T = \lambda y. f \appsep A_{xy} \appsep (B \appsep a \appsep (\lambda z.y\appsep C_z)) \appsep D_x
\end{gather}

The correct call to \elpiIn{maybe-eta} is \elpiIn{maybe-eta x T L} with
\elpiIn{L = [x]}. At first we go under the abstraction $\lambda y$ adding $y$ to
$L$. Then we find an application, where we verify that 1) $A_{xy}$ does not
contain $x$ and $y$ rigidly which is the case; 2) $B \appsep a \appsep (\lambda z.y\appsep (C_z))$
and $D_x$ can respectively reduce to $y$ and $x$. The latter reduction is
evident, since $D$ has $x$ in scope; the former subterm can reduce to $y$ since
$B$ is a variable and it exists an argument ($\lambda z.y\appsep C_z$) reducible to
$y$: under the binder $z$, we have the application of $y$ with a variable with
$z$ in scope. Note that \maybeeta, only tells if it exists a substitution making
a term an eta expansion on any term $t$, i.e, $t$ can be in $\maybeeta \cup
\maybebeta$ without the constraint of being in \llambda, as our example shows. A
possible substitution making the term $T$ in the example an $\eta$-expansion is
$\sigma = \{A \mapsto \lambda x.\lambda y.a, B \mapsto \lambda x.\lambda y.y,
C \mapsto \lambda x.x, D \mapsto \lambda x.x\}$.

\paragraph{Compilation with \linketa} Thanks to the \elpiIn{maybe-eta}
predicate, we can detect ``$\eta$-problematic'' terms and, consequently replace
them with a fresh \Ho unification variable at compilation time. The code
below illustrate this dedicated compilation:
% TODO: maybe ref to \cref{sec:nutshell}

@@

\marginpar{Where This rule is put in the code ?}
This rule is applied on \Fo{} lambda-abstractions, tests if their body $F$ is an
\maybeeta wrt to a local fresh binder $x$, and if so, it compiles $F$ to the \Ho
term \elpiIn{F1} and returns a \Ho fresh variable $A$ having in scope the free
names occurring in \elpiIn{F1}. As sketch at the very end of
\cref{sec:lang-spec}, each time a subterm $t$ in \Fo is replaced with a \Ho
variable $v$, we build a \elpiIn{link-eta} between $v$ and and the term $t'$
obtained by the compilation of $t$.

A \elpiIn{link-eta}\footnote{\elpiIn{@val-link-eta A B} is syntactic sugar for
\elpiIn{val (link-eta A B)}}
is defined as

\elpiIn{type link-eta tm -> tm -> baselink}

\noindent
We call the two terms carried by the link respectively left and right hand side.

\begin{invariant}[\linketa left hand side]
  The left hand side of a \linketa is a variable.
  \label{invariant:link-eta-left}
\end{invariant}

\begin{invariant}[\linketa right hand side]
  The right hand side of a \linketa having the shape $\lambda x.F \appsep x$
  where $F \appsep x$ is a term not starting with the \elpiIn{lam} constructor.
  \label{invariant:link-eta-right}
\end{invariant}

We want maintain invariants~\ref{invariant:link-eta-left} and \ref{invariant:link-eta-right}
